---
title: "HH612 to Submit Proj1"
output:
  html_notebook:default
---
```{r}
library(kableExtra)  ##kable
library(ModelMetrics)  ##rsme calculation
library(data.table)
library(dplyr)
library(tidyr)
library(tidyverse)
library(caTools)  ## sergio split data
```

# Purpose-Global Baseline Predictors and RMSE 
In this first assignment, we’ll attempt to predict ratings with very little information.  We’ll first look at just raw averages across all (training dataset) users.  We’ll then account for “bias” by normalizing across users and across items.   You’ll be working with ratings in a user-item matrix, where each rating may be (1) assigned to a training dataset, (2) assigned to a test dataset, or (3) missing. 


The following is an implementation of the Baseline Predictor for a Recommender System that recommends movies. This system is meant to demonstrate the usage of baseline predictors that compute biases for each user and item. The data set used is from *grouplens.org*.

# Dataset
The movie lense project was used.  The data used are extracted from the file *ml-latest-small.zip* downloaded from (http://grouplens.org/datasets/movielens/latest). 
We start by examining summaries and the first few records of the dataframes corresponding to the movies and their ratings.

```{r LoadData}
getwd()
moviefile = "C://Users//graci//Documents//ml-latest-small//movies.csv"
ratingsfile = "C://Users//graci//Documents//ml-latest-small//ratings.csv"

movies = read.csv(moviefile) ##9742 *3
ratings = read.csv(ratingsfile)  ##100836*4
head(movies,5)
head(ratings,4)
```
For the purpose of this analysis, we reduced the rows to 200
```{r ReduceSize}
summary(movies)

# Reduce the size of the data set for this assignment.
set.seed(123)
reduced = sample.int(nrow(ratings), 200, replace = F) ##int 
ratings = ratings[reduced,] ##  4
head(ratings)
summary(ratings)

# we can calculate the number of unique users as follows.
print(length(unique(ratings$userId))) ## unique users
print(length(unique(ratings$movieId)))
print(length(unique(ratings$rating)))
print(unique(ratings$rating))
```

# User-Item Matrix
A user-item matrix is created, where each row represents a user and each column a movie (item). 
The element for a given row and column represents the rating given by the user to the movie.
The data is converted from long to wide format.

```{r UserItemMatrix}
ratings2 = dplyr::select(ratings, -timestamp)  ## 3VR now, userID, movieID, rating
head(ratings2,2)

## to spread, or pivoting, 
ui_GH = spread(ratings2, movieId, rating) 
head(ui_GH,2)  
dim(ui_GH)

ui2_GH = as.matrix(dplyr::select(ui_GH, -userId))  
table(is.na(ui2_GH))
```
## Train / Test Subdata Splits
- Search for the element in the User-Item Matrix that matches the given userId, movieId, and
    - set its value to NA for the train matrix
    - set its value to the actual value for the test matrix  (not NA, !NA)

```{r SplitData}
# Randomly split all ratings for training and testing sets

split_data <- ratings2
set.seed(50)
split <- sample.split(split_data$rating, SplitRatio = 0.75)

# Prepare training set
train_data <- split_data
train_data$rating[!split] <- NA  ## this Rating col is diff from rating col
print("Training Dataset")
head(train_data,20)  ###  3VR

# Prepare testing set
test_data <- split_data
test_data$rating[split] <- NA ##  *3 VR
print("Test Dataset")    
head(test_data,20)
## aboveL train_data and test_data are mirror image of each other, same amount of obs
```
## Global Raw Rating for Entire Data 
```{r RawRating1}
raw.rating = sum(ratings$rating, na.rm=TRUE) / nrow(ratings)

print(raw.rating)
```
### RMSE for the raw average
```{r RMSE3}
library(ModelMetrics)
test_ratings_tmp$raw.rating = raw.rating

rmse(test_ratings_tmp$raw.rating, test_ratings_tmp$rating)
```
#### Global Raw Rating for Training Data / Test Data
```{r RMSE_raw2}
raw.rating.2 = sum(train_data$rating, na.rm=TRUE) / length(which(!is.na(train_data$rating)))
print(raw.rating.2)

mov_raw_avg <- round(mean(as.matrix(train_data$rating), na.rm = TRUE),4) 
mov_raw_avg  
raw_avg <- sum(train_data$rating, na.rm = TRUE) / length(which(!is.na(train_data$rating)))

# validating raw average #
print (sum(as.matrix(train_data$rating), na.rm = TRUE) / sum(!is.na(train_data$rating)) ) 

# # RMSE - Test,
 test_ratings_tmp <- subset (test_data, !is.na(test_data[,3]) ) ##3rd col is ratings, 
 test_ratings <- test_ratings_tmp [,3]##3rd col is ratings, 
  
rsq_diff <- c() 
for (i in test_ratings){  
  rsq_diff[length(rsq_diff)+1] <- (i-mov_raw_avg)^2
  }
rmse_raw_test <- sqrt(mean(rsq_diff, na.rm = TRUE))  
rmse_raw_test   
```
# Reviewers and Movie Biases
```{r Bias}

Reviewers_bias <- train_data %>% 
  filter(!is.na(rating)) %>% 
  group_by(userId) %>%
  summarise(sum = sum(rating), count = n()) %>% 
  mutate(bias = sum/count-raw_avg) %>% ## no more movieId
 
  select(userId, ReviewersBias = bias)
ReviewersBias<-Reviewers_bias$ReviewersBias

Movie_bias <- train_data %>% 
  filter(!is.na(rating)) %>% 
  group_by(movieId) %>%
  summarise(sum = sum(rating), count = n()) %>% 
  mutate(bias = sum/count-raw_avg) %>%
  select(movieId, MovieBias = bias)
MovieBias<-Movie_bias$MovieBias

train_data2 <- train_data %>% 
  left_join(Reviewers_bias, by = "userId") %>%  ##2 Reviewers Bias
  left_join(Movie_bias, by = "movieId") %>%  ##  train_data, 32 MovieBias
  mutate(RawAvg = raw_avg) %>%
  mutate(Baseline = RawAvg + ReviewersBias + MovieBias)
head(train_data2, 4)  ## * 7vr

test_data2 <- test_data %>% 
  left_join(Reviewers_bias, by = "userId") %>%
  left_join(Movie_bias, by = "movieId") %>%   ## 
  mutate(RawAvg = raw_avg) %>%
  mutate(Baseline = RawAvg + ReviewersBias + MovieBias)
 head(test_data2,2)    ## 7VR
```
``````{r RSME7}
rmse_base_train <- sqrt(sum((train_data2$rating[!is.na(train_data2$rating)] - 
                               train_data2$Baseline[!is.na(train_data2$rating)])^2) /
                           length(which(!is.na(train_data2$rating))))

rmse_base_test <- sqrt(sum((test_data2$rating[!is.na(test_data2$rating)] - 
                              test_data2$Baseline[!is.na(test_data2$rating)])^2) /
                         length(which(!is.na(test_data2$rating))))

print(rmse_base_train)  ##0.9112
print(rmse_base_test)## NA
```

```{r eval1}
UserBias <- function(uid, df) {
  ss = df[which(df$userId == uid),]
  ub = sum(ss$rating)/nrow(ss)
  return(ub)
}

ItemBias <- function(mid, df) {
  ss = df[which(df$movieId == mid),]
  ib = sum(ss$rating)/nrow(ss)
  return(ib)
}
```

# RMSE for Baseline Predictor
```{r BaselinePred}
# not working,
test_data$Baseline2 = 0.0

for (ii in 1:nrow(test_data)) {
  test_data[ii,]$Baseline2 = test_data2[ii,]$raw.rating +
                           UserBias(test_data[ii,]$userId, ratings2) - raw.rating +
                           ItemBias(test_data[ii,]$movieId, ratings2) - raw.rating
}

 rmse(test_data2$Baseline2, test_data$rating) #ERROR
```